Understanding pandoc lua filters

### 2023-09-05 supplement/bibliography/rebib.tex
More up-to date info about the texor package. 1. code: Including R code chunks. 2. rebib: Limitations of rebib 3. figure: updated the pdf and html 4. luafilters: corrected some mistakes.

\section{Summary}

>>>>

\section{Limitations}
The limitations of \CRANpkg{rebib} package are:
\begin{itemize}
\item The \CRANpkg{rebib} package only supports embedded bibliography format suggested/ commonly
used in R Journal.
\item Encoding and character set issues might not work with rebib well. Try to use UTF-8 encoding 
if possible.
\item Bibliography aggregation won't fix existing issues like missing commas/brackets.
\end{itemize}
\section{Summary}

### 2023-09-05 supplement/code-env/code-example.tex
Updated supplementary to fix some #2 issues except math and lua filters
More up-to date info about the texor package. 1. code: Including R code chunks. 2. rebib: Limitations of rebib 3. figure: updated the pdf and html 4. luafilters: corrected some mistakes.

\begin{Schunk}
Input :
\begin{Sinput}
x <- 1:100
y <- dbinom(x,100,prob = 0.5)
plot(x,y)
\end{Sinput}
Output :\\
\includegraphics[width=0.5\textwidth]{binom}
\end{Schunk}

@@

\begin{thebibliography}{2}
    \providecommand{\natexlab}[1]{#1}
    \providecommand{\url}[1]{\texttt{#1}}
    \expandafter\ifx\csname urlstyle\endcsname\relax
      \providecommand{\doi}[1]{doi: #1}\else
      \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Krewinkel, Lucero (2023)]{pandoc}
A.~ Krewinkel and A.~ Lucero
\newblock pandoc 3.0 Release notes
\newblock \emph{pandoc} \penalty0 2023
\newblock URL \url{https://pandoc.org/releases.html#pandoc-3.0-2023-01-18}

@@

\end{thebibliography}

>>>>

\begin{Schunk}
Input :
\begin{Sinput}
x <- 1:100
y <- dbinom(x,100,prob = 0.5)
plot(x,y)
\end{Sinput}
Output :\\
\includegraphics[width=1\textwidth]{binom}
\end{Schunk}

\section{Including 'R code chunks'}

Including R code chunks in an article that is to be converted to R markdown
can be a bit tricky as the \CRANpkg{knitr} package will interpret and execute the R code chunks as
actual code bits. A workaround for this issue has been described in this \href{https://support.posit.co/hc/en-us/articles/360018181633-Including-verbatim-R-code-chunks-inside-R-Markdown}{posit support article}. Below is an example with the workaround adapted from \cite{rcodechunks}.
\begin{example}
`r ''````{r}
plot(cars)
```
\end{example}

@@

\begin{thebibliography}{99}
    \providecommand{\natexlab}[1]{#1}
    \providecommand{\url}[1]{\texttt{#1}}
    \expandafter\ifx\csname urlstyle\endcsname\relax
      \providecommand{\doi}[1]{doi: #1}\else
      \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Krewinkel, Lucero (2023)]{pandoc}
A.~ Krewinkel and A.~ Lucero
\newblock pandoc 3.0 Release notes
\newblock \emph{pandoc} \penalty0 2023
\newblock URL \url{https://pandoc.org/releases.html}

@@

\bibitem[Stephens (2022)]{rcodechunks}
Nathan Stephens
\newblock Including verbatim R code chunks inside R Markdown
\newblock \emph{Posit support forum}, 2022 
\newblock URL \url{https://support.posit.co/hc/en-us/articles/360018181633-Including-verbatim-R-code-chunks-inside-R-Markdown}
\end{thebibliography}

### 2023-09-05 supplement/lua-filters/lua-filters.tex
More up-to date info about the texor package. 1. code: Including R code chunks. 2. rebib: Limitations of rebib 3. figure: updated the pdf and html 4. luafilters: corrected some mistakes.

\abstract{
pandoc supports intermediate modification of the Abstract Syntax Tree (AST) between
the parsing and writing phase using filters. This supplement paper highlights
the use cases of such filters written in the Lua language on LaTeX, markdown and native AST.
}

\section{Introduction}
To fully understand pandoc filters, one must first grasp the document conversion process. A document in one format is read/parsed into an intermediate form, known as an abstract syntax tree (AST). A document writer is a specialized type-setter to read an AST and produce the contents in the chosen article format. Converting markdown to HTML, for example, will entail these pandoc operations. Figure~\ref{fig:workflow} shows the conversion workflow within pandoc.

@@

At this point, you might wish there was a way to add such a feature. Then there is the possibility of unlimited choices for customization, including every customization as an option in pandoc or the writer is not viable. 
To address this, the idea of filters emerged, with which you could modify the AST before the writer could read it to obtain the desired result \citep{pandocfilters}.

\section{Need of Lua filters}

@@

In pandoc, there are two sorts of elements: \verb|Blocks| and \verb|Inlines|. \verb|Blocks| are complicated constructions constructed from simple pieces (\verb|Inlines|). A \verb|Para|, for example, is a \verb|Block| made up of several \verb|Str| \verb|Inlines|.

The first step in writing a filter is to select a target type, that is, which \verb|Block| or \verb|Inline| this filter will target.
Following the selection, we name the function after the target type and each element in the argument as \verb|el|.
For instance, in the case of our filter, it would be \verb|function Figure(el)|. Pandoc is smart enough to match the names of filter functions to the AST elements.

Now within the filter function we can assume that \verb|el| will contain the table object of a Figure element from the document. We can use many functions over it or \verb|Inlines| contained in it. One such function to walk over all the elements inside the block is \verb|walk_block|. Walk functions are a great way to check or count the presence of certain elements in a block. We use \verb|walk_block| to check if there are any \verb|Image| inline elements within the \verb|Figure| block. This is because after pandoc 3 \citep{pandoc} \verb|Figure| blocks can now contain elements other than \verb|Images| as well.

@@

\section{Usage of pandoc lua filters in the \CRANpkg{texor} package}
Pandoc Lua filters are used in the \CRANpkg{texor} package to modify the AST for various markups.
 Table~\ref{tab:filters} summarizes the use of each filter.
\begin{table}
\begin{tabular}{|l|p{10cm}|}
\hline
File Name & Description\\
\hline
\verb|abs_filter.lua| & {Filters out unicode \verb|182| character.}\\
\hline
\verb|bib_filter.lua| & Clears out the bibliography from the article itself as it will be added to the metadata.\\
\hline
\verb|R_code.lua| & Adds a class component \verb|'r'| to CodeBlocks for code highlighting. \\
\hline
\verb|image_filter.lua| & Fixes extensions for image paths without an extension or with \verb|.pdf| extension. \\
\hline
\verb|table_caption.lua| & Adds table numbering in the captions similar to the ones found in R Markdown and \CRANpkg{kable}-based tables. \\
\hline
\verb|image_caption.lua| & Adds Figure/Algorithm/CodeBlock/WideTable numbering as well as clears out residuals of tikz/algorithm images. \\
\hline
\verb|conversion_compat_check.lua| & This filter keeps a count of all \verb|Inline| and \verb|Block| elements and writes it to a yaml file. \\
\hline
\verb|equation_filter.lua| & Adds \CRANpkg{bookdown} style equation numbering to LaTeX equation/math environments.\\
\hline
\verb|bookdown_ref.lua| & Changes the format of links to \CRANpkg{bookdown} (markdown) style.\\
\hline
\verb|issue_checker.lua| & Searches and notifies any occurrence of unrecognized math commands.\\
\hline
\verb|find_pdf_files.lua| & This filter creates a list for all image PDF files included in the article.\\
\hline
\end{tabular}
\caption{A summary of all Lua filters used in the \CRANpkg{texor} package (available in the \href{https://github.com/Abhi-1U/texor/tree/master/inst}{/inst} folder of the package).}
\label{tab:filters}
\end{table}

>>>>

\abstract{
Pandoc supports intermediate modification of the Abstract Syntax Tree (AST) between
the parsing and writing phase using filters. This supplement paper highlights
the use cases of such filters written in the Lua language on LaTeX, markdown and native AST.
}

\section{Introduction}
To fully understand pandoc filters, one must first grasp the document conversion process.
Pandoc will read/parse an article and then store it internally as an intermediate form, known as an abstract syntax tree (AST). 
Then, A document writer(a specialized type-setter) will read the AST and produce the contents in the chosen article format. 
Converting markdown to HTML, for example, will entail these pandoc operations. 
Figure~\ref{fig:workflow} shows the conversion workflow within pandoc.

@@

At this point, you might wish there was a way to add such a feature. But including every customization as an option in pandoc or the writer is not viable. 
To address this, the idea of filters emerged, with which you could modify the AST before the writer could read it to obtain the desired result \citep{pandocfilters}.

Pandoc filters can technically be written in any language (called JSON filters) \footnote{More on this \href{https://pandoc.org/filters.html}{here}.}.
\section{Need of Lua filters}

@@

In pandoc, there are two sorts of elements: '\verb|Blocks|' and '\verb|Inlines|'. '\verb|Blocks|' are complicated constructions constructed from simpler pieces ('\verb|Inlines|'). A '\verb|Para|', for example, is a '\verb|Block|' made up of several '\verb|Str|' '\verb|Inlines|'.

The first step in writing a filter is to select a target type, that is, which '\verb|Block|' or '\verb|Inline|' this filter will target.
Following the selection, we name the function after the target type and each element in the argument as '\verb|el|'.
For instance, in the case of our filter, it would be \verb|function Figure(el)|. Pandoc is smart enough to match the names of filter functions to the AST elements.

Now within the filter function we can assume that \verb|el| will contain the table object of a Figure element from the document. We can use many functions over it or '\verb|Inlines|' contained in it. One such function to walk over all the elements inside the block is \verb|walk_block|. Walk functions are a great way to check or count the presence of certain elements in a block. We use \verb|walk_block| to check if there are any '\verb|Image|' inline elements within the '\verb|Figure|' block. This is because after pandoc 3 \citep{pandoc} '\verb|Figure|' blocks can now contain elements other than '\verb|Images|' as well.

@@

\section{Usage of pandoc lua filters in the \CRANpkg{texor} package}
Pandoc Lua filters are used in the \CRANpkg{texor} package to modify the AST for various markups.
 Table~\ref{tab:filters} summarizes the use of each filter.
\begin{table}
\begin{tabular}{|l|p{10cm}|}
\hline
File Name & Description\\
\hline
\verb|abs_filter.lua| & {Filters out unicode \verb|182| character.}\\
\hline
\verb|bib_filter.lua| & Clears out the bibliography from the article itself as it will be added to the metadata.\\
\hline
\verb|R_code.lua| & Adds a class component \verb|'r'| to CodeBlocks for code highlighting. \\
\hline
\verb|image_filter.lua| & Fixes extensions for image paths without an extension or with \verb|.pdf| extension. \\
\hline
\verb|table_caption.lua| & Adds table numbering in the captions similar to the ones found in R Markdown and \CRANpkg{kable}-based tables. \\
\hline
\verb|image_caption.lua| & Adds Figure/Algorithm numbering as well as clears out residuals of tikz/algorithm images. \\
\hline
\verb|conversion_compat_check.lua| & This filter keeps a count of all \verb|Inline| and \verb|Block| elements and writes it to a yaml file. \\
\hline
\verb|equation_filter.lua| & Adds \CRANpkg{bookdown} style equation numbering to LaTeX equation/math environments.\\
\hline
\verb|bookdown_ref.lua| & Corrects numbering for various elements.\\
\hline
\verb|issue_checker.lua| & Searches and notifies any occurrence of unrecognized math commands.\\
\hline
\verb|find_pdf_files.lua| & This filter creates a list for all image PDF files included in the article.\\
\hline
\verb|widetable_patcher.lua| & This filter sets the representation for widetables.\\
\hline
\end{tabular}

\caption{A summary of all Lua filters used in the \CRANpkg{texor} package (available in the \href{https://github.com/Abhi-1U/texor/tree/master/inst}{/inst} folder of the package).}
\label{tab:filters}
\end{table}
